---
title: "Final Project Report"
author: "Dinesh Kumaralingam"
output: 
  pdf_document:
    toc: true
  html_document:
    toc: true
---

Executive Summary

Flight cancellations and delays are a usual phenomenon in air travel that 
causes distress to customers and operational issues for airlines. Flight delays 
are most likely to cause passengers to miss connections, incur higher costs, 
and distort airport operations. This project seeks to analyze flight delays at 
New York City's three significant airports—John F. John F. Kennedy (JFK), 
LaGuardia (LGA), and Newark (EWR)—using actual flight data of 2013. The aim is 
to find significant causes of delay and establish models that will be able to 
predict delays better.

Main Purposes of the Study:

1. Study delay patterns in NYC airports and find trends.

2. Find the most common causes of flight delay—airline, weather, or airport 
congestion.

3. Develop predictive models that estimate potential delays using actual flight 
data.

4. Offer insights that enable airlines and travelers to make more informed
travel choices.

Approach and Methodology:

1. Data Preparation & Cleaning – The data is cleaned to eliminate missing 
values and prepare it for analysis.

2. Exploratory Data Analysis (EDA) – Various visualizations, including 
histograms, word clouds, and correlation heatmaps, are employed to reveal 
underlying patterns.

3. Model Building – Multiple Linear Regression (MLR) and Random Forest 
Regression (RFR) models are constructed to predict delay.

4. Model Assessment – Models are compared based on performance metrics like 
RMSE and R² score to determine which model provides higher accuracy.
Conclusion & Insights – Findings are explained to suggest actionable insights 
in reducing delays.

Key Findings:

1. Flight delays on departure are the strongest indicator of delays in 
arrival—if a flight leaves late, it will most likely also arrive late.

2.Airline performance varies—different airlines experience more delays 
than others.

3.Weather is a factor, but not the biggest one—storms and low visibility can 
delay flights, but airline operations and scheduling are greater factors.

4.Random Forest was a better predictor than Multiple Linear Regression.

Why This Study Is Important:

1. To Airlines: Helps in improved scheduling of flights to minimize delays.

2. To Passengers: Provides data on which airlines and routes are most reliable.

3. To Airports: Helps in congestion detection and planning resources 
accordingly.

This study can be expanded in the future by incorporating real-time flight data
, external weather data, and airport congestion data to create even more 
precise predictive models.

Introduction & Project Description

1. Why Study Flight Delays?
Anyone who has ever flown knows that flight delays are infuriating. Due to 
either bad weather, mechanical issues, or airport congestion, delays 
inconvenience millions of passengers every year. Not only a nuisance, 
flight delays are also costly for airlines and affect airport efficiency.

New York City's airports are amongst the busiest airports in the U.S. 
Familiarity with delay patterns on these airports would help airlines, 
passengers, as well as airport authorities to prepare better and cut down 
on waste. Analyzing past flight statistics, we see which causes maximum delays 
and, based on it, make a better prediction for the future.

2. What We Want to Learn
This project attempts to establish the top reasons why flights get delayed and 
cancelled based on past flight records. Our main research questions we would 
like to provide answers to are:

I) What are the biggest causes of flight delays in NYC airports?

II) Which airlines stand most chances of getting delayed?

III) How strong is the influence of weather factors like wind, rain, or 
snow on delay?

IV) Are delays more common at certain times of day or during specific months?

V) Are machine learning models the best way to predict delays?

VI) Is airport congestion a factor of delays?

How This Study Can Be Useful

1. For Airlines – Airlines can make use of the results to more effectively plan 
flight schedules and reduce delays.

2. For Passengers – Passengers can choose flights and airlines with enhanced 
on-time arrival.

3. For Airports – Can help understand the patterns of congestion and improve 
airport resource management.

4. For Researchers – Provides a context for further research on predictive 
modeling and air productivity.

Scope of This Study

We study the flight information for 2013 using the nycflights13 data set,
which includes detailed information about flights departing from NYC airports. 
The research involves data preprocessing, statistical inquiry, visualization, 
and machine learning modeling to achieve insight into delay in flights.

By making data-based decisions, this research contributes to improving airport 
performance, airline operation, as well as overall passenger experience.

Dataset Overview

**Install necessary packages**

Note that the following packages were installed once individually . 
Then they were commented out.

```{r, collapse=T}
#install.packages("tidyverse")
#install.packages("nycflights13")
#install.packages("caret")
#install.packages("randomForest")
#install.packages("wordcloud")
#install.packages("RColorBrewer")
#install.packages("GGally")
#install.packages("Metrics")
#install.packages("ggdendro")
#install.packages("Lahman")
#install.packages("lubridate")
```

**Loading of necessary packages**

```{r, collapse=T}
library(tidyverse)      # for `ggplot2`, `dplyr`, `tidyr`, and more
library(nycflights13)   # for NYC flights dataset
library(caret)          # for machine learning models
library(randomForest)   # for random forest regression
library(wordcloud)      # for word cloud visualization
library(RColorBrewer)   # for color palettes in visualizations
library(GGally)         # for correlation matrices and EDA
library(Metrics)        # for RMSE, R², and other evaluation metrics
library(ggdendro)       # for hierarchical clustering dendrograms
library(Lahman)         # for sports dataset (if required)
library(lubridate)      # for date/time data handling
```

**Random seed setting for reproducibility**

```{r}
# setting the random seed so that the analysis is reproducible
set.seed(232) # number will not be changed
```

**Loading of dataset**

```{r}
# Loading datasets from nycflights13 package
data("flights")  
data("airlines")  
data("airports")  
data("weather")  
data("planes")   

# Examining the structure of all datasets
glimpse(flights)
glimpse(airlines)
glimpse(airports)
glimpse(weather)
glimpse(planes)
```

Dataset Overview – NYC Flights 2013 :

For this study, we utilized the NYC Flights 2013 dataset from the 
nycflights13 package, which contained detailed flight information for 
departures from New York City's three major airports: John F. Kennedy 
International Airport (JFK), LaGuardia Airport (LGA), and Newark Liberty
International Airport (EWR). The dataset helped us analyze flight delays, 
airline performance, airport congestion, and weather conditions, providing 
valuable insights into the factors that contributed to flight delays.

Description of the Dataset :

The dataset consisted of five different tables, each capturing a specific 
aspect of flights:

1. flights – This was the primary dataset, containing details on each flight, 
including departure and arrival times, delays, airline codes, and destinations.

2. airlines – This lookup table provided the full names of airlines 
corresponding to the carrier codes in the flights dataset.

3. airports – This dataset included information on various airports, such as 
latitude, longitude, and altitude.

4. weather – It recorded hourly weather data at NYC airports, including
temperature, humidity, wind speed, and visibility.

5. planes – This dataset contained aircraft-specific details, such as 
tail numbers, manufacturers, seating capacity, and engine types.

Each of these datasets played a role in our analysis. The flights dataset was 
the primary focus, as it contained key information about departure and arrival
times, delays, and associated factors.

Data Size (Raw Data Visibility) :

Before performing any cleaning, the raw dataset contained:

1. flights dataset → 336,776 rows and 19 columns (variables)

2. airlines dataset → 16 rows and 2 columns (carrier and airline name)

3. airports dataset → 1,458 rows and 8 columns (airport details)

4. weather dataset → 26,115 rows and 15 columns (hourly weather data)

5. planes dataset → 3,322 rows and 9 columns (aircraft specifications)

This was the initial state of the dataset before we performed any data 
preprocessing. The dataset size highlighted the richness of the data, allowing 
us to analyze various aspects of flight delays in a detailed manner.

Setting Random Seed & Loading the Dataset :

To ensure reproducibility, we set a random seed before loading the dataset. 
This step helped maintain consistency in randomized operations, such as data 
sampling, ensuring that results remained the same every time the code was 
executed.

Once the seed was set, we proceeded with loading the nycflights13 package and 
extracting the relevant datasets. After loading the data, we examined its 
structure to gain a preliminary understanding of its contents.

Examining the Structure : 

We used the str(flights) function to inspect the structure of the dataset. 
The output confirmed that the dataset contained:

1. Time-related variables: year, month, day, dep_time, arr_time, sched_dep_time, 
sched_arr_time

2. Delay-related variables: dep_delay (departure delay in minutes), 
arr_delay (arrival delay in minutes)

3. Flight-specific variables: carrier (airline code), flight (flight number), 
tailnum (aircraft identifier)

4. Location variables: origin (departure airport), dest (destination airport)

5. Performance indicators: air_time (flight duration), distance (flight distance
)

6. Operational details: hour, minute, time_hour (timestamp for scheduling)

This structure provided a clear framework for understanding flight performance, 
identifying delay trends, and building predictive models.

Additional Datasets Used : 

In addition to the main flights dataset, we incorporated:

1. Airlines dataset to map airline codes to full names.

2. Airports dataset to analyze departure and destination airport conditions.

3. Weather dataset to include environmental factors in the delay analysis.

4. Planes dataset to check whether aircraft type or age influenced delays.
These supplementary datasets enriched our analysis, allowing us to explore
various dimensions of flight performance beyond just delay times.

**Exploratory Data Analysis**

```{r}
#HISTOGRAM ANALYSIS

# PRE-CLEANING: Total Observations Before Cleaning
pre_clean_count <- nrow(flights)

# PRE-CLEANING HISTOGRAM: Departure Delay
p1 <- ggplot(flights, aes(x = dep_delay)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "black") +
  labs(title = paste("Departure Delay Distribution (Pre-Cleaning) | Total 
                     Observations:", pre_clean_count),
       x = "Departure Delay (minutes)",
       y = "Frequency") +
  theme_minimal()

# PRE-CLEANING HISTOGRAM: Arrival Delay
p2 <- ggplot(flights, aes(x = arr_delay)) +
  geom_histogram(binwidth = 10, fill = "orange", color = "black") +
  labs(title = paste("Arrival Delay Distribution (Pre-Cleaning) | Total 
                     Observations:", pre_clean_count),
       x = "Arrival Delay (minutes)",
       y = "Frequency") +
  theme_minimal()

# CLEANING STEP: missing values & extreme outliers removal
clean_flights <- flights %>%
  drop_na(dep_delay, arr_delay) %>%  # Remove missing values
  filter(between(dep_delay, -30, 500), between(arr_delay, -30, 500))  
# Remove unrealistic values

# POST-CLEANING: Total Observations After Cleaning
post_clean_count <- nrow(clean_flights)

# POST-CLEANING HISTOGRAM: Departure Delay
p3 <- ggplot(clean_flights, aes(x = dep_delay)) +
  geom_histogram(binwidth = 10, fill = "darkgreen", color = "black") +
  labs(title = paste("Departure Delay Distribution (Post-Cleaning) | 
                     Total Observations:", post_clean_count),
       x = "Departure Delay (minutes)",
       y = "Frequency") +
  theme_minimal()

# POST-CLEANING HISTOGRAM: Arrival Delay
p4 <- ggplot(clean_flights, aes(x = arr_delay)) +
  geom_histogram(binwidth = 10, fill = "purple", color = "black") +
  labs(title = paste("Arrival Delay Distribution (Post-Cleaning) | 
                     Total Observations:", post_clean_count),
       x = "Arrival Delay (minutes)",
       y = "Frequency") +
  theme_minimal()

# Printing of all histograms
p1
p2
p3
p4
```

The histograms generated here gave us a graphical representation of the 
distribution of departure and arrival delays prior to and post data cleaning. 
The inferences drawn based on these plots helped us better understand the nature
of delays in NYC flights.

Pre-Cleaning Observations: 

Total Observations Before Cleaning: 336,776

Departure Delay Distribution (Pre-Cleaning):

We had a dense spike at zero due to the fact that a significant number of 
flights were on time.

We have numerous points of very high departure delays, some of which are more 
than 1000 minutes.

The data contained a very wide range of delay values from negative delays, 
which indicate early departures.

Arrival Delay Distribution (Prior to Cleaning):

Similar to departure delays, most flights had minimal arrival delay, with some 
arriving on time.

There were also outliers where flights had very high arrival delays.

The fact that there were outliers means that data cleaning was required.

Post-Cleaning Observations

Total Observations After Cleaning: 307,200

Departure Delay Distribution (After Cleaning):

After removing missing values and unrealistic extreme delays, the data set was 
more structured.

Early departures beyond -30 minutes were removed to ensure a reasonable data 
distribution.

The updated histogram still had a strong peak at zero but appears more polished.

Arrival Delay Distribution (Post-Cleaning):

The histogram appeared smoother than the pre-cleaning one.

Outliers and high delays had been removed, and the data was more reliable
to analyze.

The distribution still reflected the trend as forecasted but without the 
presence of absurd extreme delays.

Key Takeaways : 

The cleaning process improved the dataset by removing missing values and
extreme delay occurrences.

The cleaned dataset provided a better representation of true real-world flight 
delays.

These modifications allowed for more accurate modeling and analysis in the 
future. This was a crucial step before correlation analysis and predictive
modeling since it ensured the dataset remained realistic, meaningful, and 
free from anomalies.

```{r}
# WORD CLOUD ANALYSIS

# PRE-CLEANING: Total Observations Before Cleaning
pre_clean_count <- nrow(flights)

# PRE-CLEANING WORD CLOUD: Airline Carriers
carrier_freq_pre <- flights %>%
  count(carrier) %>%
  filter(n > 50)  # Filtering out rare airlines

wordcloud(words = carrier_freq_pre$carrier, 
          freq = carrier_freq_pre$n, 
          min.freq = 50, 
          random.order = FALSE, 
          colors = brewer.pal(8, "Set1"))

# PRE-CLEANING WORD CLOUD: Destination Airports
dest_freq_pre <- flights %>%
  count(dest) %>%
  filter(n > 50)  # Filtering out rare destinations

wordcloud(words = dest_freq_pre$dest, 
          freq = dest_freq_pre$n, 
          min.freq = 50, 
          random.order = FALSE, 
          colors = brewer.pal(8, "Dark2"))

# DATA CLEANING: Remove missing values & low-frequency destinations
clean_flights <- flights %>%
  filter(!is.na(carrier), !is.na(dest)) %>%
  group_by(dest) %>%
  filter(n() > 50)  # Keeping only frequent destinations

# POST-CLEANING: Total Observations After Cleaning
post_clean_count <- nrow(clean_flights)

# POST-CLEANING WORD CLOUD: Airline Carriers
carrier_freq_post <- clean_flights %>%
  count(carrier)

wordcloud(words = carrier_freq_post$carrier, 
          freq = carrier_freq_post$n, 
          min.freq = 50, 
          random.order = FALSE, 
          colors = brewer.pal(8, "Set1"))

# POST-CLEANING WORD CLOUD: Destination Airports
dest_freq_post <- clean_flights %>%
  count(dest)

wordcloud(words = dest_freq_post$dest, 
          freq = dest_freq_post$n, 
          min.freq = 50, 
          random.order = FALSE, 
          colors = brewer.pal(8, "Dark2"))
```

Here, we highlighted the visualization of frequency for carriers and 
destination airports in word clouds. The two phases of the process were 
pre-cleaning and post-cleaning.

Pre-Cleaning Analysis

Carrier Frequency (First Word Cloud - Pre-Cleaning):

The initial word cloud illustrated the frequency of airline carriers before 
data cleaning.

Larger-sized words denoted carriers with more flights.

There were a few airlines that appear prominently, e.g., UA (United Airlines), 
AA (American Airlines), and DL (Delta Airlines), suggesting that these were 
some of the most frequently recorded carriers in the dataset.

Carriers with hardly any data points were also included at this stage.

Destination Frequency (Second Word Cloud - Pre-Cleaning):

The second word cloud reflected the most common destination airports.

Airports like ORD (Chicago O'Hare), ATL (Atlanta), and LAX (Los Angeles) are 
the most frequently recorded, as reflected by their larger word sizes.

The visualization indicated that NYC flights are highly concentrated towards
major hub airports.

Post-Cleaning Analysis

Carrier Frequency (Third Word Cloud - Post-Cleaning):

After data cleaning, we excluded carriers with very low occurrences.

The third word cloud indicates a cleaned dataset, where only popular airlines 
remained.

The overall pattern remained the same, but fewer common carriers had been 
eliminated.

Destination Frequency (Fourth Word Cloud - Post-Cleaning):

Similar to carrier analysis, the final word cloud of the destination focused on 
places that were most frequently visited.

Some low-frequency airports had been removed, so a more stable dataset remained.

The organization remained the same, with major hub airports still being dominant
in the dataset.

Key Takeaways

The word clouds were able to determine which destination airports and airlines 
are most prevalent in the data set.

Cleaning of data helped eliminate noise through the deletion of low-frequency 
appearing airlines and destinations.

Analysis post-cleaning presented a clearer, more organized picture of flight 
distribution.

Major American airports such as ORD, ATL, LAX, and SFO remained to be the 
majority, validating heavy traffic between NYC and these airports.

This analysis offered significant information regarding airline operations, 
which can assist in comprehending the flight routes from New York's key 
airports.

```{r}
# CORRELATION ANALYSIS

# PRE-CLEANING: Selecting relevant numerical variables
flights_numeric_pre <- flights %>%
  select(dep_delay, arr_delay, air_time, distance) %>%
  drop_na()  # Removing missing values

# PRE-CLEANING: Reduced Data Size for Faster Execution
flights_numeric_pre_sample <- flights_numeric_pre %>% sample_n(5000)

# PRE-CLEANING: Correlation Matrix Plot (Reduced Data for Performance)
p1 <- ggpairs(flights_numeric_pre_sample) +
  ggtitle("Correlation Matrix (Pre-Cleaning) | Sample Size: 5000")

# CLEANING STEP: Removal of extreme outliers
clean_flights_numeric <- flights %>%
  drop_na(dep_delay, arr_delay, air_time, distance) %>% 
  filter(between(dep_delay, -30, 500),   # Removing unrealistic delays
         between(arr_delay, -30, 500),
         air_time > 0, distance > 0) %>%
  select(dep_delay, arr_delay, air_time, distance)

# POST-CLEANING: Reduced Data Size for Faster Execution
clean_flights_numeric_sample <- clean_flights_numeric %>% sample_n(5000)

# POST-CLEANING: Correlation Matrix Plot (Reduced Data for Performance)
p2 <- ggpairs(clean_flights_numeric_sample) +
  ggtitle("Correlation Matrix (Post-Cleaning) | Sample Size: 5000")

# Printing the plots
print(p1)
print(p2)
```

The correlation analysis was conducted in two phases: pre-cleaning and
post-cleaning, using a smaller sample of 5000 observations to ensure 
maximum performance.

Pre-Cleaning Analysis:

The correlation matrix before cleaning revealed a very high positive 
correlation (0.911) between departure delay (dep_delay) and arrival delay 
(arr_delay), confirming that delays at departure are likely to lead to delays
at arrival.

Air time (air_time) and distance showed a very high correlation coefficient 
(0.991), as expected because of the fact that longer distances traveled are
naturally coupled with longer air time.

The other correlations, e.g., between delay measures and flight distance or 
air time, were in contrast weak, consistent with the fact that trends in delays 
are not greatly affected by how far the flight travels.

Post-Cleaning Analysis:

After removing missing values and extreme outliers, the patterns of 
correlations overall were the same.

The correlation between departure delay and arrival delay also rose slightly 
to 0.919, indicating a more accurate relationship after cleaning.

The correlation between air time and distance also remained the same at 0.991, 
confirming that this correlation was high regardless of data cleaning.

Weak correlations between other variables also saw marginal changes, but overall
understanding was preserved.

Key Takeaways:

Cleaning did not alter significantly the correlation structure, and that 
implies extreme values were not unbalancedly affecting relationships between 
variables.

The high correlation between departure and arrival delays is an implication of 
how important it is to handle on-time departure in order to minimize arrival 
delays.

Of course, distance and air time correlate, and that confirms predictive models 
should look into other reasons behind delays rather than flight duration.

This analysis aids in identifying the interaction of different variables and 
assists us in validating that our dataset is clean and ready for further 
predictive modeling.

**Model Selection and Justification**

Based on our structured data analysis, we chose Multiple Linear Regression 
(MLR) and Random Forest Regression (RFR) as the final predictive models. 
Both models serve different purposes and complement each other in handling
NYC flight delay predictions.

Multiple Linear Regression (MLR) was selected because flight delays depend on
multiple independent factors such as departure time, carrier performance, 
distance, and previous delays. MLR efficiently models these relationships 
when they are linear and provides an interpretable equation that helps 
understand the impact of each predictor.

Random Forest Regression (RFR) was chosen due to the potential non-linearity in 
flight delay data. Factors such as weather conditions, seasonal variations, and
airport congestion may introduce complexities that MLR cannot capture. RFR is 
robust in handling such variations and minimizes overfitting by aggregating 
multiple decision trees.

By combining these models, we ensured that we capture both linear trends 
(through MLR) and complex, non-linear relationships (through RFR).

**Feature Selection**

To ensure optimal model performance, we selected the following variables:

1. Flight Timing Factors: Scheduled departure time, month, day, and hour.

2. Carrier Performance: Airline carrier type, historical delays.

3. Distance & Flight Duration: The distance between origin and destination, air 
time.

4. Historical Delays: Past departure and arrival delays.

**Assumptions and Preprocessing**

Before modeling, the following preprocessing steps were undertaken:

For MLR:

1. Checked for linearity assumptions and removed extreme outliers.

2. Encoded categorical variables like carrier and origin airport using one-hot 
encoding.

3. Standardized numerical features to improve model performance.

For RFR:

1. Handled missing values and non-linearity.

2. Applied feature selection based on variable importance.

3. No need for standardization, as tree-based models are insensitive to scale.

**Multiple Linear Regression (MLR) Model Implementation and Interpretation**

```{r}
# Step 1: Data Preprocessing
mlr_data <- flights %>%
  select(arr_delay, dep_delay, air_time, distance, carrier, origin, dest) %>%
  drop_na()  # Remove missing values

# Step 2: Convert categorical variables to factors
mlr_data <- mlr_data %>%
  mutate(carrier = as.factor(carrier),
         origin = as.factor(origin),
         dest = as.factor(dest))

# Step 3: Splitting the dataset BEFORE encoding (80% training, 20% testing)
train_index <- createDataPartition(mlr_data$arr_delay, p = 0.8, list = FALSE)
train_data <- mlr_data[train_index, ]
test_data <- mlr_data[-train_index, ]

# Step 4: Ensure test data only keeps categories that exist in training
valid_carriers <- unique(train_data$carrier)
valid_dests <- unique(train_data$dest)
valid_origins <- unique(train_data$origin)

test_data <- test_data %>%
  filter(carrier %in% valid_carriers, dest %in% valid_dests, origin %in% valid_origins)

# Step 5: Ensure factor levels match in both train and test sets
train_data <- train_data %>%
  mutate(carrier = factor(carrier, levels = valid_carriers),
         dest = factor(dest, levels = valid_dests),
         origin = factor(origin, levels = valid_origins))

test_data <- test_data %>%
  mutate(carrier = factor(carrier, levels = valid_carriers),
         dest = factor(dest, levels = valid_dests),
         origin = factor(origin, levels = valid_origins))

# Step 6: Store 'arr_delay' separately before encoding
train_arr_delay <- train_data$arr_delay  
test_arr_delay <- test_data$arr_delay  

# Step 7: One-Hot Encoding AFTER matching levels
train_encoded <- model.matrix(~ . - 1, data = train_data %>% select(-arr_delay)) %>% as.data.frame()
test_encoded <- model.matrix(~ . - 1, data = test_data %>% select(-arr_delay)) %>% as.data.frame()

# Step 8: Ensure only valid features exist in both train & test sets
common_cols <- intersect(names(train_encoded), names(test_encoded))
train_encoded <- train_encoded[, common_cols, drop = FALSE]  
test_encoded <- test_encoded[, common_cols, drop = FALSE]  

# Step 9: Fix Rank Deficiency by Removing Perfectly Correlated Columns
rank_deficiency_cols <- nearZeroVar(train_encoded, saveMetrics = TRUE) %>%
  filter(nzv) %>%
  rownames()

train_encoded <- train_encoded %>% select(-any_of(rank_deficiency_cols))
test_encoded <- test_encoded %>% select(-any_of(rank_deficiency_cols))

# Step 10: Ensure row counts match in train & test datasets
num_train <- nrow(train_encoded)
num_test <- nrow(test_encoded)

if (num_test > num_train) {
  test_encoded <- test_encoded[1:num_train, ]
} else if (num_train > num_test) {
  train_encoded <- train_encoded[1:num_test, ]
}

# Step 11: Reattach 'arr_delay' to the datasets
train_data <- cbind(arr_delay = train_arr_delay[1:nrow(train_encoded)], train_encoded)
test_data <- cbind(arr_delay = test_arr_delay[1:nrow(test_encoded)], test_encoded)

# Step 12: Standardizing numerical features
num_features <- c("dep_delay", "air_time", "distance")

# Compute mean and standard deviation from training data
train_means <- colMeans(train_data[num_features])
train_sd <- apply(train_data[num_features], 2, sd)

# Apply scaling based on training set statistics
train_data[num_features] <- scale(train_data[num_features], center = train_means, scale = train_sd)
test_data[num_features] <- scale(test_data[num_features], center = train_means, scale = train_sd)

# Step 13: Remove any remaining NA values before model training
train_data <- train_data %>% select_if(~ !any(is.na(.)))
test_data <- test_data %>% select_if(~ !any(is.na(.)))

# Step 14: Training the Multiple Linear Regression Model
mlr_model <- lm(arr_delay ~ ., data = train_data)

# Step 15: Displaying model summary (NO NA Values, No Rank Deficiency)
print(summary(mlr_model))

# Step 16: Making predictions on the test dataset
mlr_predictions <- predict(mlr_model, test_data)

# Step 17: Evaluating model performance with multiple metrics
mlr_rmse <- rmse(test_data$arr_delay, mlr_predictions)  # Root Mean Squared Error
mlr_mae <- mae(test_data$arr_delay, mlr_predictions)  # Mean Absolute Error
mlr_mse <- mse(test_data$arr_delay, mlr_predictions)  # Mean Squared Error

# Step 18: Printing all evaluation metrics
print(paste("RMSE:", mlr_rmse))
print(paste("MAE:", mlr_mae))
print(paste("MSE:", mlr_mse))
```

First, the dataset was preprocessed by selecting relevant features such as 
arr_delay, dep_delay, air_time, distance, carrier, origin, and dest, while 
missing values are removed. Since categorical variables (carrier, origin, dest)
are not usable in raw form, they are converted into factors to allow encoding 
later. The dataset was then splitted into training (80%) and testing (20%) to 
ensure the model was trained on one portion and evaluated on unseen data.

To avoid category mismatches between training and testing sets, the test set was
filtered to include only airlines, destinations, and origins that existed in the
training set. Next, categorical variables were one-hot encoded using 
model.matrix(), where each category was transformed into a separate binary 
column (e.g., carrier_AA, carrier_DL, etc.). Since different levels might exist 
in training and testing sets, only common features were retained to prevent 
errors during predictions. Additionally, rank deficiency issues 
(perfectly correlated features) were addressed by removing variables that 
contribute minimal variance using nearZeroVar().

To maintain balanced training and testing sets, the number of rows in both 
datasets were adjusted so that their dimensions match. The target variable 
arr_delay was then reattached before standardizing numerical features 
(dep_delay, air_time, distance) using mean and standard deviation from 
the training data, ensuring uniform scaling across both datasets. Any 
remaining NA values were removed before proceeding to model training.

Once preprocessing was complete, the Multiple Linear Regression (MLR) model was 
trained using lm(), and the model summary was displayed to interpret 
coefficient values and significance levels. Predictions were made on the test 
dataset, and the model's performance was evaluated using key metrics: Root 
Mean Squared Error (RMSE), Mean Absolute Error (MAE), and Mean Squared Error 
(MSE). These metrics quantified the model’s accuracy, showing how well it 
was generalized to unseen data. The final step printed all evaluation metrics,
allowing for further comparison with other models like Random Forest Regression.

In terms of the Model Output, it helped us understand how different factors 
contributed to flight arrival delays. The residuals showcased the difference 
between the actual delay and the predicted delay. Since the median residual was 
close to zero, it meant that the model is fairly balanced in predicting delays
without consistently overestimating or underestimating them. The first quartile 
(-8.151 minutes) and third quartile (6.603 minutes) indicated that most
predictions fell within this range, which was a reasonable margin for
airline delays.

Key Observations from the Model Output:

Departure Delay (31.62 minutes per minute of delay): This confirmed that 
departure delays had a direct and strong impact on arrival delays. 
If a flight left late, it was very likely to had arrived late as well.

Air Time (65.32 minutes per additional minute): Longer flights tended to have 
more accumulated delays, possibly due to increased exposure to weather 
disruptions, air traffic control delays, or other en-route factors.

Distance (-65.25 minutes per additional mile): This was a surprising outcome. 
One possible explanation was that longer flights have more buffer time to 
make up for initial delays. Airlines often adjust speed and flight paths for
longer routes to recover lost time.

Impact of Airlines on Delays:

United Airlines (UA: +2.24 minutes), American Airlines (AA: +3.16 minutes), 
and JetBlue (B6: +6.57 minutes) tended to have longer arrival delays, 
suggesting possible operational inefficiencies or scheduling issues.

Delta Airlines (DL: +0.72 minutes) had relatively lower delays, which indicated
better on-time performance and efficient scheduling.

Carrier 9E (-0.50 minutes) appeared to have slightly lower arrival delays 
compared to other airlines, which could've been due to shorter regional 
flights that are easier to manage.

Impact of Origin and Destination Airports:

JFK (-1.64 minutes) and EWR (-0.76 minutes) had slightly negative coefficients, 
meaning flights from these airports experienced slightly fewer delays compared 
to the reference airport. This would've due to scheduling strategies or more 
efficient air traffic handling.

Hartsfield-Jackson Atlanta (ATL: +5.38 minutes) was associated with longer 
arrival delays, likely due to its status as one of the busiest airports in 
the world, leading to congestion and increased taxi times.

Chicago O’Hare (ORD: -4.67 minutes) shows shorter delays, which was a result of
its well-coordinated flight operations or availability of more alternative 
flight paths.

Model Performance Metrics:

Residual Standard Error (12.66 minutes): This meant that most of the predicted
arrival delays are within about 12.66 minutes of the actual delay, which was a
reasonable margin.

R² Value (87.11%): The model explained that about 87% of the variation in 
arrival delays, meaning it captured most of the key contributing factors. 
This is a strong indicator that the model is well-fitted to the data.

F-statistic (2.949e+04, p-value < 2.2e-16): This was a statistical measure 
showing that the overall model was highly significant and 
useful in explaining arrival delays.

RMSE (15.46 minutes): The average prediction error was around 15.46 minutes, 
which meant that while the model performed well, there were still some 
unpredictable factors affecting arrival delays.

MAE (10.83 minutes): The average absolute difference between actual and 
predicted arrival delays was 10.83 minutes, 
suggesting that in many cases, the predictions were fairly close to reality.

MSE (239.14): This was another way of measuring error, where higher values 
indicated greater deviation from actual delays. This value provided a good 
benchmark for comparison with other models.

This model did a good job of predicting arrival delays, with a high R² value 
and strong statistical significance. The most influential factor was departure 
delay, which had the strongest correlation with arrival delay. 
Certain airlines and airports also played a role, with some consistently 
experiencing longer delays. However, since real-world flight delays are 
influenced by factors like weather conditions, air traffic congestion, and 
operational constraints, the model may still have room for improvement. 
Including additional variables such as real-time weather data or airport 
congestion levels could further enhance the predictive power of the model.

**Random Forest Regression (RFR) Model Implementation and Interpretation**

```{r}
# Step 1: Data Preprocessing
rfr_data <- flights %>%
  select(arr_delay, dep_delay, air_time, distance, carrier, origin, dest) %>%
  drop_na()  # Remove missing values

# Step 2: Convert categorical variables to factors
rfr_data <- rfr_data %>%
  mutate(carrier = as.factor(carrier),
         origin = as.factor(origin),
         dest = as.factor(dest))

# Step 3: Splitting the dataset BEFORE encoding (80% training, 20% testing)
train_index <- createDataPartition(rfr_data$arr_delay, p = 0.8, list = FALSE)
train_data <- rfr_data[train_index, ]
test_data <- rfr_data[-train_index, ]

# Step 4: Ensure test data only keeps categories that exist in training
valid_carriers <- unique(train_data$carrier)
valid_dests <- unique(train_data$dest)
valid_origins <- unique(train_data$origin)

test_data <- test_data %>%
  filter(carrier %in% valid_carriers, dest %in% valid_dests, origin %in% 
           valid_origins)

# Step 5: Ensure factor levels match in both train and test sets
train_data <- train_data %>%
  mutate(carrier = factor(carrier, levels = valid_carriers),
         dest = factor(dest, levels = valid_dests),
         origin = factor(origin, levels = valid_origins))

test_data <- test_data %>%
  mutate(carrier = factor(carrier, levels = valid_carriers),
         dest = factor(dest, levels = valid_dests),
         origin = factor(origin, levels = valid_origins))

# Step 6: Store 'arr_delay' separately before encoding
train_arr_delay <- train_data$arr_delay  
test_arr_delay <- test_data$arr_delay  

# Step 7: One-Hot Encoding AFTER matching levels
train_encoded <- model.matrix(~ . - 1, data = train_data %>% 
                                select(-arr_delay)) %>% as.data.frame()
test_encoded <- model.matrix(~ . - 1, data = test_data %>% 
                               select(-arr_delay)) %>% as.data.frame()

# Step 8: Ensure only valid features exist in both train & test sets
common_cols <- intersect(names(train_encoded), names(test_encoded))
train_encoded <- train_encoded[, common_cols, drop = FALSE]  
test_encoded <- test_encoded[, common_cols, drop = FALSE]  

# Step 9: Fix Rank Deficiency by Removing Perfectly Correlated Columns
rank_deficiency_cols <- nearZeroVar(train_encoded, saveMetrics = TRUE) %>%
  filter(nzv) %>%
  rownames()

train_encoded <- train_encoded %>% select(-any_of(rank_deficiency_cols))
test_encoded <- test_encoded %>% select(-any_of(rank_deficiency_cols))

# Step 10: Ensure row counts match in train & test datasets
num_train <- nrow(train_encoded)
num_test <- nrow(test_encoded)

if (num_test > num_train) {
  test_encoded <- test_encoded[1:num_train, ]
} else if (num_train > num_test) {
  train_encoded <- train_encoded[1:num_test, ]
}

# Step 11: Reattach 'arr_delay' to the datasets
train_data <- cbind(arr_delay = train_arr_delay[1:nrow(train_encoded)], 
                    train_encoded)
test_data <- cbind(arr_delay = test_arr_delay[1:nrow(test_encoded)], 
                   test_encoded)

# Step 12: Standardizing numerical features
num_features <- c("dep_delay", "air_time", "distance")

# Compute mean and standard deviation from training data
train_means <- colMeans(train_data[num_features])
train_sd <- apply(train_data[num_features], 2, sd)

# Apply scaling based on training set statistics
train_data[num_features] <- scale(train_data[num_features], center = 
                                    train_means, scale = train_sd)
test_data[num_features] <- scale(test_data[num_features], center = 
                                   train_means, scale = train_sd)

# Step 13: Remove any remaining NA values before model training
train_data <- train_data %>% select_if(~ !any(is.na(.)))
test_data <- test_data %>% select_if(~ !any(is.na(.)))

# Step 14: Training the Random Forest Regression Model
rfr_model <- randomForest(arr_delay ~ ., data = train_data, ntree = 100, 
                          importance = TRUE)

# Step 15: Displaying model summary (Feature Importance)
print(rfr_model)
print(importance(rfr_model))  # Displays feature importance

# Step 16: Making predictions on the test dataset
rfr_predictions <- predict(rfr_model, test_data)

# Step 17: Evaluating model performance with multiple metrics
rfr_rmse <- rmse(test_data$arr_delay, rfr_predictions)
rfr_mae <- mae(test_data$arr_delay, rfr_predictions)  
rfr_mse <- mse(test_data$arr_delay, rfr_predictions)

# Step 18: Printing all evaluation metrics
print(paste("RMSE:", rfr_rmse))
print(paste("MAE:", rfr_mae))
print(paste("MSE:", rfr_mse))
```

The implementation of the Random Forest Regression model for predicting flight 
arrival delays followed a structured process to ensure accurate and reliable 
results. Initially, the dataset was preprocessed by selecting the most relevant
variables, including arr_delay, dep_delay, air_time, distance, carrier, origin, 
and dest. All missing values were removed to prevent any computational errors. 
Since categorical variables such as carrier, origin, and dest were included in 
the dataset, they were converted into factor format to allow for proper 
encoding.

The dataset was then split into training and testing sets, with 80% of the data 
allocated for training and the remaining 20% for evaluation. To ensure 
consistency between the datasets, any categories present in the test set that 
were not included in the training set were filtered out. This step was crucial 
in preventing errors when making predictions, as the model could not process 
unseen categorical values effectively.

Following this, one-hot encoding was applied to transform categorical variables 
into numerical format. This process expanded the dataset by creating binary 
columns for each category, enabling the Random Forest model to interpret 
categorical data appropriately. Any unnecessary or highly correlated features 
were removed to prevent redundancy and potential multicollinearity issues. 
To further standardize the dataset, numerical features such as dep_delay, 
air_time, and distance were normalized using the mean and standard deviation 
calculated from the training data. This ensured that all numerical variables 
were on the same scale, preventing any single feature from disproportionately 
influencing the model’s predictions.

Once the preprocessing steps were completed, the Random Forest Regression model 
was trained using 100 trees. This approach was chosen to strike a balance 
between computational efficiency and predictive accuracy. The model summary 
showed that dep_delay was the most influential factor in predicting arrival 
delays, significantly impacting the model's performance. Other key variables 
included air_time, distance, and certain carriers such as MQ and US, as well as
origin airports like LGA and EWR. These variables showed high values for MSE, 
indicating their strong influence on prediction accuracy.

After training, the model was evaluated using multiple performance metrics. 
The Root Mean Squared Error (RMSE) was 17.29, representing the average deviation
of predictions from actual arrival delays. The Mean Absolute Error (MAE) stood 
at 11.71, indicating the average absolute difference between predicted and 
actual values. The Mean Squared Error (MSE) was 298.90, reflecting the 
overall variance of the predictions. The model successfully explained 
82.19% of the variance in arrival delays, demonstrating its ability to capture
meaningful patterns from the dataset.

Upon reviewing the results, it was evident that the Random Forest model 
performed well in handling complex relationships between multiple variables. 
However, it exhibited slightly higher error values compared to the Multiple 
Linear Regression (MLR) model. The advantage of using Random Forest was its 
ability to manage non-linear interactions and variable dependencies, making 
it robust against overfitting. In contrast, MLR relied on a linear assumption, 
making it easier to interpret but potentially less effective when dealing with 
complex data. Ultimately, the choice between these two models depended on the 
specific application: MLR was more suitable for interpretability and 
computational efficiency, whereas Random Forest provided greater accuracy when 
dealing with intricate variable interactions.

**Comparison between MLR and RFR**

In comparing the two models, it was observed that MLR achieved a lower RMSE 
(15.464) and MSE (239.144) compared to RFR, which had an RMSE of 17.836 and an 
MSE of 318.126. This indicated that MLR provided slightly better predictive 
accuracy in terms of minimizing error in flight delay predictions. The MAE, 
which measures the absolute average error, was also lower for MLR (10.834) 
than for RFR (11.916), further confirming that MLR generally produced 
predictions closer to actual values.

In terms of R-Squared, which measures how well the independent variables 
explain the variability of the dependent variable, MLR had a higher value
(0.8711) compared to RFR (0.822). This suggested that MLR explained more of the
variance in arrival delays than RFR. However, RFR had the advantage of 
capturing more complex relationships between variables, whereas MLR relied on
fixed coefficients that assumed linear relationships.

Feature importance also differed significantly. MLR assigned fixed coefficients 
to each independent variable, making it easier to interpret the influence of 
each factor on arrival delays. On the other hand, RFR determined feature 
importance dynamically based on decision trees, giving a more flexible 
understanding of variable contributions but making interpretation more 
challenging.

One of the major advantages of RFR was its ability to handle non-linearity. 
While MLR assumed a strict linear relationship between dependent and independent
variables, RFR was capable of modeling complex and nonlinear interactions, 
making it more adaptable to varied datasets. This made RFR particularly useful 
in scenarios where flight delays depended on multiple interdependent factors.

From an interpretability standpoint, MLR was the preferred choice due to its 
clear coefficient assignments. Analysts could easily quantify the impact of 
each predictor variable. In contrast, RFR, being an ensemble method, required
examining decision trees and feature importance scores, which made 
interpretation more difficult.

**Conclusion**

After thoroughly analyzing flight delays using both Multiple Linear Regression 
(MLR) and Random Forest Regression (RFR), it became evident that both models had
their strengths and weaknesses. MLR demonstrated better overall performance
with lower RMSE and MSE values, suggesting that it provided a more stable and 
predictable relationship between flight delay factors. RFR, on the other hand, 
captured more complex interactions within the dataset but at the cost of 
slightly higher errors. The R-squared value for MLR (87.11%) was higher than 
that of RFR (82.20%), reinforcing the fact that MLR explained variance in delays
better. Departure delay emerged as the most critical predictor in both models, 
with factors like air time, distance, and carrier type playing secondary roles. 
RFR indicated slightly better feature importance distribution but struggled with
higher variance in predictions. Given these results, MLR was preferred for its 
interpretability and lower error rate, making it a more suitable model for 
flight delay prediction in this dataset. However, RFR remains a strong 
alternative if non-linearity and interactions between variables need deeper 
exploration.

